\section{Introduction}

Dynamic routing is a coupled sequential decision problem rather than a
collection of independent shortest-path queries.  At each decision interval,
every active connected and autonomous vehicle (CAV) chooses one valid route
from a masked set of up to \(K\) candidates connecting its current road segment
to its destination.  When several CAVs favor the same route, their decisions
reshape congestion and the
travel conditions faced by later vehicles.  Mixed-autonomy traffic makes this
coupling asymmetric: the controller can coordinate CAVs, but human-driven
vehicles (HDVs) remain self-interested and cannot be rerouted directly.  The
goal is therefore to coordinate the controllable vehicles so as to reduce the
global mean travel time of both populations in a dynamic routing game
\cite{calderone2017markov,lazar2020routing}.

Analytical approaches characterize such congestion-coupled routing through
traffic assignment, equilibrium, mean-field, or leader--follower models
\cite{tanaka2020linearly,cabannes2022solving,kolarich2022stackelberg}.
These formulations provide interpretable descriptions of traffic externalities
and human response.  Their application to dynamic mixed-autonomy control,
however, commonly requires prescribed link-cost functions, behavioral
assumptions, or repeated equilibrium computation.  These requirements become
restrictive when time-varying departures, incidents, and microscopic vehicle
interactions jointly determine the traffic transition.

Model-free multi-agent reinforcement learning (MARL) offers a complementary
route: a microscopic simulator supplies the transition dynamics, and CAVs
learn state-dependent route policies from experience.  Prior work formulates
dynamic routing as a Markov game and provides executable route-choice
environments for urban networks
\cite{lazar2021learning,shou2022multi,akman2025routerl,akman2026urb}.
This flexibility replaces an explicit analytical model with simulator-generated
training transitions.  It does not, however, remove the cost of repeatedly
querying the traffic environment.  Moreover, a route decision may affect a
shared bottleneck only after several decision intervals, which makes real
interaction difficult to reuse for long-horizon, multi-vehicle credit
assignment.

World-model reinforcement learning suggests a way to reuse these interactions:
learn predictive latent dynamics from collected trajectories, then perform
additional policy updates in imagined rollouts
\cite{hafnerdream,hafnermastering,hafner2025mastering}.  Applying this idea to
mixed-autonomy routing raises three coupled challenges.  First, traffic evolves
as a graph-structured spatiotemporal field, while a route index has meaning only
within one vehicle's current candidate set; many local route choices must
therefore be converted into a graph-level transition condition.  Second, HDVs
form an uncontrolled background population whose aggregate route demand can
vary across episodes and must be estimated from preceding CAV pressure and HDV
demand.  Third, actions are taken by individual vehicles, whereas the objective
is network-wide, so policy learning must preserve vehicle-specific credit
without ignoring global congestion effects.  Driving and multi-agent world
models demonstrate the value of latent imagination
\cite{mamba,li2024think2drive,yang2026raw2drive}.  In the
mixed-autonomy routing formulation studied here, however, jointly handling
these interfaces remains an open problem.

We propose UrbanDreamer, a world-model-based MARL framework designed around
these interfaces.  A parameter-shared actor retains the executable control
semantics: each active CAV selects a current-to-destination candidate route.
UrbanDreamer projects the sampled route choices onto a planned CAV pressure
field defined on road-segment nodes, placing vehicle actions and network
dynamics in the same graph coordinate system.  A graph convolutional encoder
represents the traffic field, and a state-to-state flow-matching model predicts
its next latent representation conditioned on CAV pressure, departures,
episode context, and a history-conditioned estimate of HDV route demand.
During imagined rollouts, a deterministic vehicle wrapper maintains identities,
route progress, departures, and arrivals, while centralized per-vehicle and
global critics
provide individual and network-level value signals.  Simulator collection
episodes refresh the world model, while held-out simulator episodes remain the
basis for policy evaluation.

The technical contributions are threefold:
\begin{itemize}
  \item We formulate a world-model-based MARL framework for mixed-autonomy
  dynamic routing whose goal is to reduce real simulator interaction:
  additional CAV policy updates are performed in learned latent traffic
  dynamics instead of repeated microscopic simulation.
  \item We introduce a graph-aligned action interface and traffic world model
  that connect executable per-CAV route choices to segment-level latent
  transitions, combining graph convolutional traffic encoding with conditioned
  state-to-state flow matching, together with a history-conditioned,
  slow-timescale model of aggregate HDV response.
  \item We develop identity-preserving imagined route learning with a
  parameter-shared CAV actor, a centralized per-vehicle critic, and an auxiliary
  global critic.  A two-stage, simulator-grounded training pipeline alternates
  online traffic collection, world-model adaptation, and imagined actor--critic
  optimization.
\end{itemize}

Experiments on three urban road networks instantiated through the RouteRL and
Urban Routing Benchmark (URB) framework, containing 636, 1,035, and 2,068
scheduled vehicles, show that UrbanDreamer reaches strong routing policies
with substantially fewer real simulator interactions and lower wall-clock
training time than model-free MARL baselines, and attains the lowest global
mean travel time on two of the three networks.
