\section{Preliminaries}

\subsection{Problem Formulation}

In this section, we formally define the mixed-autonomy dynamic routing problem
studied in this paper. A typical scenario contains connected and autonomous vehicles‌(CAVs) and
human-driven vehicles(HDVs) moving over a directed road network under
time-varying departure demand. The road network, vehicle population, route
candidates, segment-indexed traffic fields, and decision timing constitute the
basic components of the problem. We first define these components and then cast
the control process as a Markov decision process.
% 中文注释：本节正式定义本文研究的 mixed-autonomy dynamic routing problem。一个典型场景包含在有向路网和随时间变化的出发需求下行驶的可控 CAVs 与自利且不可控的 HDVs。road network、vehicle population、route candidates、segment-indexed traffic fields 和 decision timing 构成该问题的基本组成部分。我们先定义这些组成部分，然后将控制过程表述为一个 Markov decision process。

\noindent\textbf{Definition 1 (Segment-Node Road Network).}
We represent the road network as a directed graph
\(G=(N,E)\).  Each node \(u\in N\) is a road segment on which traffic
state can be measured, predicted, and controlled through route choices.  A
directed edge \((u,v)\in E\) exists when vehicles can move directly from road
segment \(u\) to road segment \(v\).
% 中文注释：我们将 road network 表示为一个 directed graph \(G=(N,E)\)。每个 node \(u\in N\) 是一条 road segment，其上的 traffic state 可以被测量、预测，并通过 route choices 受到控制。当车辆可以从 road segment \(u\) 直接移动到 road segment \(v\) 时，存在 directed edge \((u,v)\in E\)。

Each segment node has static attributes such as length, speed limit, lane
count, road type, capacity, and free-flow travel time. A fixed node ordering assigns every road segment a stable index and stores the adjacency matrix \(A\).
% 中文注释：每个 segment node 具有 length、speed limit、lane count、road type、capacity 和 free-flow travel time 等静态属性。固定的 node ordering 为每条 road segment 分配稳定 index，并存储 \(G\) 的 row-normalized adjacency matrix \(A\in\mathbb{R}^{M\times M}\)。


\noindent\textbf{Definition 2 (Vehicles).}
We partition the departure horizon, spanning from the beginning to the final scheduled vehicle departure, into fixed decision intervals indexed by decision epochs \(t=0,\ldots,T\). At decision epoch \(t\), the active vehicle population is
\(\mathcal{N}_t=\mathcal{N}^{\mathrm{CAV}}_t\cup
\mathcal{N}^{\mathrm{HDV}}_t\).  CAVs are controlled by the learned routing
policy, whereas HDVs are self-interested agents whose routing behavior can not be controlled in this problem. Each vehicle \(i\) has its origin segment
\(o_i\), destination segment \(g_i\), a scheduled departure time
\(\tau_i^{\mathrm{dep}}\), and a type
\(m_i\in\{\mathrm{CAV},\mathrm{HDV}\}\). If vehicle \(i\) is active at
decision epoch \(t\), \(u_{i,t}\) denotes its current segment node.
% 中文注释：在 decision epoch \(t\)，active vehicle population 为 \(\mathcal{N}_t=\mathcal{N}^{\mathrm{CAV}}_t\cup \mathcal{N}^{\mathrm{HDV}}_t\)。CAVs 由学习得到的 routing policy 控制，而 HDVs 是自利的 agents，其 routing behavior 不由 UrbanDreamer 直接控制。每辆车 \(i\) 具有 origin segment \(o_i\)、destination segment \(g_i\)、scheduled departure time \(\tau_i^{\mathrm{dep}}\)，以及 type \(m_i\in\{\mathrm{CAV},\mathrm{HDV}\}\)。如果车辆 \(i\) 在 decision epoch \(t\) 处于 active 状态，则 \(u_{i,t}\) 表示其 current segment node。


\noindent\textbf{Definition 3 (Route Candidates and Decision Timing).}
When CAV \(i\) is active at decision epoch \(t\), it must select one route from
route candidates
\begin{equation}
\mathcal{P}_{i,t}
=
\{P^1_{i,t},\ldots,P^{K}_{i,t}\}.
\end{equation}
Here each candidate route is a sequence of segment nodes
connecting the current segment \(u_{i,t}\) to the destination segment \(g_i\).
The CAV action is the route index \(a_{i,t}\in\{1,\ldots,K\}\), which
selects candidate route \(P^{a_{i,t}}_{i,t}\). In contrast, an HDV makes its
route choice only at its departure time \(\tau_i^{\mathrm{dep}}\)(considering human travel intent).
% 中文注释：当 CAV \(i\) 在 decision epoch \(t\) 处于 active 状态时，它必须从有限的 current-to-destination candidate set \(\mathcal{P}_{i,t}=\{P^1_{i,t},\ldots,P^{K_i(t)}_{i,t}\}\) 中选择一条路线。这里 \(K_i(t)\le K\)，每条 candidate path 是一个 segment nodes 序列，将 current segment \(u_{i,t}\) 连接到 destination segment \(g_i\)。CAV action 是 route index \(a_{i,t}\in\{1,\ldots,K_i(t)\}\)，它选择 candidate path \(P^{a_{i,t}}_{i,t}\)。相反，HDV 只在其 departure time \(\tau_i^{\mathrm{dep}}\) 做 route choice（考虑到人类出行意图）

Route candidates are generated by the Dijkstra algorithm on the road segment
graph. Let \(\ell(u)\) and \(v_{\max}(u)\) denote the length and speed limit of
road segment \(u\). The initial segment cost is its free-flow travel time,
\begin{equation}
\begin{array}{l}
c^1(u)=t_{\mathrm{ff}}(u)=
\ell(u)/v_{\max}(u).
\end{array}
\end{equation}
For \(k=1,\ldots,K\), the route generator attempts Dijkstra under the current cost
\(c^k\). After a route is found, all segment nodes on that route are penalized
before the next search:
\begin{equation}
\begin{array}{rcl}
P^k_{i,t}
&=&
\mathrm{Dijkstra}(G,u_{i,t},g_i;c^k),\\
c^{k+1}(u)
&=&
\left\{
\begin{array}{ll}
\eta c^k(u), & u\in P^k_{i,t},\\
c^k(u), & u\notin P^k_{i,t},
\end{array}
\right.
\end{array}
\end{equation}
where \(\eta>1\) is the penalty. The search stops when no new
valid route can be found.
% 中文注释：candidate routes 由 segment graph 上的 repeated Dijkstra search 生成。令 \(\ell(u)\) 和 \(v_{\max}(u)\) 表示 segment node \(u\) 的 length 和 speed limit。初始 segment cost 是它的 free-flow travel time，满足 \(c^1(u)=t_{\mathrm{ff}}(u)\)，且 \(t_{\mathrm{ff}}(u)=\ell(u)/\max(v_{\max}(u),\varepsilon)\)。对于 \(k=1,\ldots,K\)，generator 会尝试在当前 cost \(c^k\) 下运行 Dijkstra。找到一条 route 之后，该 route 上的所有 segment nodes 会在下一次搜索之前被惩罚：\(R^k_{i,t}=\mathrm{Dijkstra}(G,u_{i,t},g_i;c^k)\)，且如果 \(u\in R^k_{i,t}\)，则 \(c^{k+1}(u)=\eta c^k(u)\)，否则 \(c^{k+1}(u)=c^k(u)\)。这里 \(\eta>1\) 是 route-diversity penalty。当找不到新的有效 route 时，搜索停止。



\subsection{Markov Decision Process}

Mixed-autonomy routing game is a sequential decision-making problem over decision epochs. We model the route-choice process as a Markov
decision process \(\mathcal{M}=(\mathcal{S},\mathcal{A},\mathcal{T},r,\gamma)\),
defined as follows.
% 中文注释：mixed-autonomy routing 是一个随固定 decision epochs 展开的 sequential decision-making problem。我们将 route-choice process 建模为 finite-horizon Markov decision process \(\mathcal{M}=(\mathcal{S},\mathcal{A},\mathcal{T},r,\gamma)\)，定义如下。

\noindent\textbf{Definition 4 (Observation/State).}
At decision epoch \(t\), the global state \(s_t\in\mathcal{S}\) includes the
road graph \(G\), scheduled vehicles with their attributes, and current route candidates. Each active agent(CAV) observes a
projection of the global state \(o_{i,t}=\mathcal{O}_i(s_t)\), which
contains the current position, destination, candidate routes, and route-level
traffic summaries.

We also define a segment graph observation projected from the global state \(X_t=\chi(s_t,G)\). For each road segment \(u\), \(X_t(u)\) contains both static and dynamic features:
\begin{equation}
\begin{array}{l}
X_t(u)=
[\mathrm{c}_t(u),\ v_t(u),
\kappa(u),
\ell(u),\ v_{\max}(u),
\lambda(u)].
\end{array}
\end{equation}
Here \(\mathrm{c}_t\) is vehicle counts, \(v_t\) is dynamic speed, \(\kappa\) is capacity, \(\ell\) is
length, and \(v_{\max}\) is speed limit, \(\lambda\) is lane count.

The tensor \(X_t\) is not equivalent to \(s_t\).  It is a lossy
segment-aligned view that removes vehicle identities, individual destinations,
and chosen route suffixes.  These components re-enter the model through the
vehicle observations \(o_{i,t}\), the planned CAV pressure, the departure
demand, the HDV influence, and the episode context, which together condition
the learned transition.

% 中文注释：在 decision epoch \(t\)，global state \(s_t\in\mathcal{S}\) 包括 road graph \(G\)、current time、带有 types、origins、destinations 和 departure times 的 scheduled vehicles、active CAV 和 HDV 集合、active vehicles 的 current segment 和 route status、episode 中更早的 route choices、当前 CAV candidate-route lists，以及 departure demand \(D_t\)。每个 agent 观察一个投影 \(o_{i,t}=\mathcal{O}_i(s_t)\)。对于 active CAV，\(o_{i,t}\) 包含其 current segment、destination、candidate routes、route-level traffic summaries、route-cost estimates 和 validity information；对于 HDV，该 observation 是 background human-routing behavior 使用的 departure-time information。



\noindent\textbf{Definition 5 (Action).}
Let
\(\mathcal{I}_t\subseteq\mathcal{N}^{\mathrm{CAV}}_t\) denote the active CAVs
that can be routed at decision epoch \(t\). The joint action is
\begin{equation}
\begin{array}{rcl}
a_t&=&(a_{i,t})_{i\in\mathcal{I}_t},\\
\mathcal{A}_t&=&
\displaystyle\prod_{i\in\mathcal{I}_t}\{1,\ldots,K\}.
\end{array}
\end{equation}
where \(a_{i,t}=k\) selects candidate route \(P^k_{i,t}\). HDV route choices
are generated by the background human-routing behavior at departure time and
are treated as part of the environment.
% 中文注释：agent set 是 scheduled vehicle population，但只有 CAVs 是可控的。令 \(\mathcal{I}_t\subseteq\mathcal{N}^{\mathrm{CAV}}_t\) 表示 decision epoch \(t\) 中可以被 routing 的 active CAVs。joint action 为 \(a_t=(a_{i,t})_{i\in\mathcal{I}_t}\)，\(\mathcal{A}_t=\prod_{i\in\mathcal{I}_t}\{1,\ldots,K\}\)。其中 \(a_{i,t}=k\) 选择 candidate route \(P^k_{i,t}\)。HDV route choices 由 background human-routing behavior 在 departure time 生成，并被视为 environment 的一部分。

\noindent\textbf{Definition 6 (Transition).}
Given the current state, joint CAV action, departure demand\(D_t\)during the decision interval, and HDV traffic state(influence caused by HDV routing, such as speed and position)
, the transition is
\begin{equation}
\begin{array}{l}
s_{t+1}\sim
\mathcal{T}(\cdot\mid s_t,a_t,D_t,H_t^{\mathrm{HDV}}).
\end{array}
\end{equation}
With newly departing vehicles, HDV traffic state, selected CAV routes, the transition progresses vehicles and updates road segments' speed and occupancy, and active sets of vehicles.
% 中文注释：给定 current state、joint CAV action、departure demand 和 HDV background behavior，transition 为 \(s_{t+1}\sim\mathcal{T}(\cdot\mid s_t,a_t,D_t,H_t^{\mathrm{HDV}})\)。该 transition 注入新出发车辆，记录 HDV departure-time route choices，应用 selected CAV routes，使车辆在 segment-node graph 上前进，并更新 queues、speeds、arrivals 和 active sets。

\noindent\textbf{Definition 7 (Objective/Reward).}
At the end of an episode, each completed vehicle(both CAV and HDV) has travel time
\(\Delta_i=t_i^{\mathrm{arr}}-\tau_i^{\mathrm{dep}}\). The main optimization objective is system-level: minimize the expected global mean travel time over all completed vehicles,
\begin{equation}
\begin{array}{rcl}
\min\quad
\mathcal{T}_{\mathrm{all}}&=&
\displaystyle
{E}_{\pi_\theta,P}
\left[
\frac{1}{|\mathcal{N}^{\mathrm{arr}}|}
\sum_{i\in\mathcal{N}^{\mathrm{arr}}}
\Delta_i
\right].
\end{array}
\end{equation}

Here \(\mathcal{N}^{\mathrm{arr}}\) is the set of arrived vehicles, 
\(t_i^{\mathrm{arr}}\) is the arrival time of vehicle \(i\), and \(\tau_i^{\mathrm{dep}}\) is the departure time of vehicle \(i\).

%这里有可能需要补充step或者vehicle级别的reward， 如果method部分决定要写这里就不用补了

% 中文注释：episode 结束时，每辆完成行程的车辆都有 travel time \(\Delta_i=t_i^{\mathrm{arr}}-\tau_i^{\mathrm{dep}}\)。主要目标是 Definition~2 中的 system-level mean travel time。per-step reward \(r_t=r(s_t,a_t,s_{t+1})\) 是与更低 travel time、更低 route cost 或更快 route progress 对齐的 learning signal，policy return 为 \(J_{\mathrm{RL}}(\pi_\theta)=\mathbb{E}_{\pi_\theta,\mathcal{T}}[\sum_{t=0}^{T-1}\gamma^t r_t]\)。evaluation 将 \(\mathcal{T}_{\mathrm{all}}\) 作为主要指标报告，并将分开的 CAV 和 HDV mean travel times 作为诊断指标。


