\section{Training Pipeline}

UrbanDreamer follows a two-stage training pipeline.  The first stage provides warmup learning for the dynamics model. The second stage
alternates real simulator collection, causal HDV-response learning, online
world-model updates, and imagined actor-critic optimization after every
collected episode. Consequently, the model is continually adapted to the
state-action distribution induced by the improving CAV policy, while policy
learning remains grounded in newly observed traffic transitions.
% 中文注释：UrbanDreamer 采用两阶段训练流程。第一阶段为 traffic world model 提供稳定的初始化。第二阶段在每个 collected episode 之后，依次交替进行 real simulator collection、causal HDV-response learning、online world-model updates 和 imagined actor-critic optimization。因此，world model 会持续适应不断改进的 CAV policy 所诱导的 state-action distribution，而 policy learning 仍以新观测到的 traffic transitions 为基础。

\subsubsection{Offline Warmup}

Before online policy learning, we collect a fixed set of transitions from the simulator using exploratory CAV route choices. The encoder, flow-matching dynamics, and decoder are
then fitted on these transitions using \(\mathcal{L}_\phi\). 
% 中文注释：在 online policy learning 之前，我们从 post-mutation simulator state 收集一组固定 transitions；该阶段使用 exploratory CAV route choices，并冻结 HDV adaptation。随后，encoder、flow-matching dynamics 和 decoder 使用 \(\mathcal{L}_\phi\) 在这些 transitions 上完成初始化训练。causal HDV response predictor 不在该阶段训练，因为固定 collection 不能表示由持续变化的 CAV policy 引起的 episode-to-episode response。actor、critics 和 response predictor 在逐 episode online stage 开始时初始化。

\subsubsection{Online Training}

During online collection, HDV adaptation is enabled and the current actor
stochastically samples a valid route for every active CAV.  The same sampled
one-hot decision is used to reroute the vehicle, construct
\(\Psi_{e,t}^{\mathrm{CAV}}\), and store the world-model action in replay.
This action alignment ensures that each observed transition is paired with the
physical CAV intervention that actually produced it.  Deterministic masked
argmax is reserved for simulator evaluation rather than training collection.
% 中文注释：在 online collection 中，HDV adaptation 被启用，current actor 为每辆 active CAV 随机采样一条有效 route。同一个 sampled one-hot decision 同时用于对车辆执行 rerouting、构造 \(\Psi_{e,t}^{\mathrm{CAV}}\)，以及在 replay 中保存 world-model action。这样的 action alignment 保证每个 observed transition 都与实际产生它的物理 CAV intervention 配对。deterministic masked argmax 仅用于 simulator evaluation，而不用于 training collection。

Only episodes satisfying the replay acceptance criterion are admitted to the
bounded world-model replay buffer \(\mathcal{B}\).  Before the observed target
\(H_e^\star\) of an accepted episode is exposed to the response predictor,
\(\hat H_e^{\mathrm{HDV}}\) is computed using only summaries from preceding
episodes and is attached to that episode's transitions.  After
\(H_e^\star\) is recorded, the predictor is updated on the available adjacent
episode pairs using \(\mathcal{L}_\psi\), and the world model is independently
updated on real transition minibatches from \(\mathcal{B}\) using
\(\mathcal{L}_\phi\).  This ordering prevents the current HDV target from
leaking into its own causal prediction.
% 中文注释：只有满足 replay acceptance criterion 的 episodes 才会进入有界的 world-model replay buffer \(\mathcal{B}\)。在 accepted episode 的 observed target \(H_e^\star\) 暴露给 response predictor 之前，\(\hat H_e^{\mathrm{HDV}}\) 只根据先前 episodes 的 summaries 计算，并附加到当前 episode 的 transitions 上。记录 \(H_e^\star\) 后，predictor 使用 \(\mathcal{L}_\psi\) 在已有的 adjacent episode pairs 上更新；world model 则使用 \(\mathcal{L}_\phi\) 在从 \(\mathcal{B}\) 采样的 real transition minibatches 上独立更新。这个顺序防止 current HDV target 泄漏到它自身的 causal prediction 中。

The main online configuration does not apply a direct actor update to real
simulator returns.  Instead, valid CAV decision states are sampled from
\(\mathcal{B}\) to initialize the vehicle-level imagined rollouts described
above.  Within each rollout, the actor samples masked route actions, the same
one-hot actions define the vehicle wrapper and graph-level CAV pressure, and
the world model predicts the resulting traffic evolution.  The resulting
per-vehicle and global lambda returns jointly update the actor and both
critics.
% 中文注释：当前 online 主线不直接使用 real simulator returns 更新 actor。相反，训练从 \(\mathcal{B}\) 中采样包含有效 CAV decisions 的 states，并用它们初始化前文介绍的 vehicle-level imagined rollouts。在每个 rollout 内，actor 采样 masked route actions；相同的 one-hot actions 同时用于 vehicle wrapper 和 graph-level CAV pressure；world model 则预测由这些 actions 导致的 traffic evolution。最终得到的 per-vehicle 和 global lambda returns 共同用于更新 actor 和两个 critics。

\begin{algorithm}[t]
\caption{Online training}
\label{alg:urbandreamer}
\small
\begin{algorithmic}
\REQUIRE Road graph \(G\), simulator, training-episode budget, checkpoint indices
\ENSURE Simulator-selected CAV route actor \(\pi_\theta\)
\FOR{each online episode \(e\)}
  \STATE Collect one simulator episode using masked stochastic route samples from \(\pi_\theta\)
  \STATE Reuse each sampled one-hot route for execution, \(\Psi_{e,t}^{\mathrm{CAV}}\), and replay
  \IF{the episode satisfies the replay acceptance criterion}
    \STATE Predict \(\hat H_e^{\mathrm{HDV}}\) from preceding episode summaries
    \STATE Attach \(\hat H_e^{\mathrm{HDV}}\), append the episode to \(\mathcal{B}\), and discard stale transitions
    \STATE Update \(g_\psi^{\mathrm{HDV}}\) on available adjacent episode pairs
  \ENDIF
  \IF{a world-model update is scheduled and \(\mathcal{B}\) is nonempty}
    \FOR{each configured world-model update}
      \STATE Sample real transitions from \(\mathcal{B}\) and minimize \(\mathcal{L}_\phi\)
    \ENDFOR
  \ENDIF
  \IF{\(\mathcal{B}\) contains enough valid CAV decision states}
    \FOR{each configured imagined actor-critic update}
      \STATE Sample replay start states and initialize their vehicle states
      \FOR{each of the next \(L_{\mathrm{im}}\) imagined decisions}
        \STATE Sample masked routes from \(\pi_\theta\) and construct \(\Psi_t^{\mathrm{CAV}}\)
        \STATE Predict \(Z_{t+1}\) using the stored causal HDV context
        \STATE Advance the vehicle wrapper and record rewards and active masks
      \ENDFOR
      \STATE Compute lambda returns and update the actor and both critics
    \ENDFOR
  \ENDIF
  \IF{\(e\) is a checkpoint episode}
    \STATE Save the actor, critics, world model, and response predictor
  \ENDIF
\ENDFOR
\STATE Evaluate saved actors in the simulator using masked argmax routes
\RETURN The actor with the lowest global mean travel time
\end{algorithmic}
\end{algorithm}
% 中文注释：Algorithm~\ref{alg:urbandreamer} 首先在冻结 HDV adaptation 的固定 warm-up transitions 上初始化 world model。进入 online stage 后，每个 episode 都使用 current actor 的 masked stochastic route samples 进行 simulator collection，并将同一个 sampled one-hot route 用于车辆执行、planned CAV pressure 和 replay。满足 acceptance criterion 的 episode 会先得到仅依赖先前 episode summaries 的 causal HDV prediction，再进入 bounded replay；随后分别更新 HDV response predictor 和 world model。当 replay 中存在足够的有效 CAV decision states 时，算法从 replay 初始化 imagined rollouts，在每一步采样 route、推进 world model 和 vehicle wrapper，并根据 lambda returns 联合更新 actor 与两个 critics。训练完成后，所有 saved actors 使用 masked argmax 在 simulator 中评估，并按照 global mean travel time 选择最终 actor。

At designated episodes, UrbanDreamer saves the actor, critics, world model,
and response predictor together.  After online training, each saved actor is
evaluated in the real simulator with masked argmax route selection, and the
checkpoint with the lowest global mean travel time is selected.  Simulator
selection prevents favorable world-model errors from determining the final
policy solely through imagined returns.
% 中文注释：在指定 episodes，UrbanDreamer 会同时保存 actor、critics、world model 和 response predictor。online training 完成后，每个 saved actor 都使用 masked argmax route selection 在真实 simulator 中评估，并选择 global mean travel time 最低的 checkpoint。基于 simulator 的选择可以防止有利于 policy 的 world-model errors 仅通过 imagined returns 决定最终 policy。
